Death By Dropdown?

### 2025-04-10 conference-materials/2025_shiny-conf/slides.qmd
Adds link to slides in GitHub repository
- Updates slide presentations with a link to the GitHub repository where the slides are hosted, improving accessibility for users.
Merge ShinyConf 2025 Materials & Set Up BID Framework Repository (#1)
* Adds Elsevier journal template and BID framework outline

- Introduces a Quarto extension for generating Elsevier journal
articles, including citation styles and templates.
- Integrates BID framework outlines for application, software, and theory papers, providing structured content for each manuscript type.
- Enhances user experience by aligning with El

---
title: "Death By Dropdown?"
subtitle: "A Developer's Guide To Building Dashboards That Won't Fry Your Client's Brain"
author: "Milena Eickhoff & Jeremy Winget, PhD"
format: 
  revealjs:
    title-slide-attributes:
      data-background-image: "img/title.png"
      data-background-size: "100% 100%"
    theme: simple
    slide-number: false
    footer: "ShinyConf 2025 | Death By Dropdown?"
    incremental: true
    transition: slide
    background-transition: fade
    highlight-style: github
    code-block-background: true
---

@@

::: {.notes}
- Have you ever felt like this? You were asked to develop a dashboard with endless dropdown menus, overwhelming complecity, that you know would leave users disengaged and searching for answers? 
-	Send us an emoji if you have felt like this. How many of you have even abandoned a dashboard because it was too complex?
:::

@@

- Our framework flows through 5 stages: Notice → Interpret → Structure → Anticipate → Validate
  * Click through
- Notice identifies friction points where users struggle with the interface
- Interpret uncovers what story users truly need - moving from symptoms to diagnosis
- Structure organizes information in ways that align with how our brains process information
- Anticipate considers cognitive biases affecting data interpretation
- Validate ensures users leave with clear takeaways and a positive impression
- Remember, these stages are iterative - you'll often revisit earlier stages as you discover new insights
:::

@@

::: {.column .fragment }
- **UX Design**: Established best practices
- **Data Storytelling**: From data dumps to meaningful narratives
- **Psychological Science**: Evidence-based cognitive principles
- **Visual Communication**: Optimized for information perception
:::

@@

- BID integrates established best practices from multiple disciplines - we're truly standing on the shoulders of giants
  * Click through example fields...UX, data storytelling, psych science, data/visual communication
- What makes BID uniquely powerful is how it systematically connects psychological principles to specific coding decisions at each development stage
- Unlike typical checklists, BID guides how you think about dashboards from psychological first principles
- The framework bridges theory and practice - showing when and how to apply principles in your workflow
- You don't need a psychology degree to apply these effectively
- Let's see how these stages transformed SoundPulse's dashboard
:::

@@

- **Stage 1 (Notice):** Users were overwhelmed with options
  * This relates to 'cognitive load' - the mental effort needed to process information
  * When the brain gets overtaxed, comprehension drops
  * Hick's Law shows decision time increases logarithmically with more choices
- **Stage 2 (Interpret):** We identified what users needed - increased Processing Fluency
  * When information is easy to process, people find it more credible
  * We focused on the key story executives needed, not showing all possible data
- **Stage 3 (Structure):** We applied the 'Principle of Proximity' in the structure stage...which is about grouping related elements together
  * We also make use of Dual-Processing Theory (which distinguishes between fast, intuitive thinking and slower, deliberative thinking) in the design to create logical groupings of metrics for more intuitive workflows
- Now let's see how we brought these concepts to life
:::

@@

```{r}
#| eval: false
#| echo: true
#| code-line-numbers: "4-5|6-18|19-23"

@@

```{r}
#| eval: false
#| echo: true
#| code-line-numbers: "2-10|12-16|17-27|29-38|39|40-56"

@@

- In the more advanced example we will show you how you can group columns and create a dynamic legend.
  * First, we create a vector that stores the colors we want to use to distinguish the subgroups. 
  * Then we create the reactable. You can use the columnGroups   option to group primary and secondary metrics together (example of the principle of proximity) – it creates a header above the columns. 
  * Below we create a container for the dynamic legend and map over the colors and age groups using a function from the purrr package. The dynamic legend reduces choice complexity for the user.
- Ask: “How many have used columnGroups in reactable before?” (engagement opportunity) 
  * If not many…great, we taught something!
  * If a lot…could also mention (quickly) how to extend with legend/div

@@

- For the later BID stages, we addressed subtler psychological factors affecting interpretation and action
- **Stage 4 (Anticipate):** We expected some folks to make flawed comparisons without proper context
  * This relates to the 'Anchoring Effect'...where people rely too heavily on the first piece of information they see
  * We also saw teams struggling to interpret the data in the same way, which connects to 'Framing & Loss Aversion'...how the presentation of identical data as a "gain" or a "loss" significantly affects decisions
- **Stage 5 (Validate):** Finally, we applied the 'Peak-End Rule'...the idea that people judge experiences by their most intense moment and ending
  * This helped us understand why users needed clear, memorable takeaways to conclude their dashboard experience
- These psychological principles guided our approach to the remaining implementation
:::

@@

```{r}
#| eval: false
#| echo: true
#| code-line-numbers: "2-21|23-30"

@@

```{r}
#| eval: false
#| echo: true
#| code-line-numbers: "2-21|23-29"

@@

```{r}
#| eval: false
#| echo: true
#| code-line-numbers: "2-3|6-12|14-20|23-24|25-40|42-43|45-59"

@@

  # Grab KPI column names
  kpi_col_names <- c("popularity", "danceability", "energy")

  # Generate all chart outputs dynamically
  walk(kpi_col_names, function(kpi_col) {
    output_id <- paste0("chart_", kpi_col)

@@

- For the framing toggle we use radio buttons that trigger the server-side code to dynamically recalculate the values based on the framing perspective. Different teams at SoundPulse used different frames - marketing used Progress, product used Gap. This is psychological design directly impacting business decisions . 
  * In the last part we loop over the different kpi columns (using a function from the purrr package) and update all charts dynamically with the function from the previous page.
:::

@@

- We're not just sharing theories - we're giving you practical tools for immediate implementation
- The `{bidux}` package directly implements our 5-stage framework
- Each function maps to a specific stage and guides your thinking
- Phase 1 is available today on GitHub with the conceptual foundation
- Future phases will include LLM integration for suggesting implementations
- We're also developing pre-built UI components embodying these principles
- Think of `{bidux}` as your behavioral science assistant for Shiny development
:::

@@

::: {.fragment}
```{r}
#| echo: true
library(bidux)

@@

notice_stage
notice_stage$theory
notice_stage$suggestions
```
:::

@@

::: {.r-stack}
::: {.fragment .fade-out}
```{r}
#| echo: true

# Concept dictionary with fuzzy searching and case-insensitive lookup
bid_concept("processing fluency")
```
:::

::: {.fragment .fade-in-then-out}
```{r}
#| echo: true

# Concept dictionary with fuzzy searching and case-insensitive lookup
bid_concept("processing fluency")$description
bid_concept("processing fluency")$implementation_tips
```
:::

::: {.fragment .fade-in-then-out}
```{r}
#| echo: true

# Concept dictionary with fuzzy searching and case-insensitive lookup
cog_concepts <- bid_concepts("cog")
paste("Found", nrow(cog_concepts), "cog-related concepts")
```
:::

::: {.fragment}
```{r}
#| echo: true

# BID Notice stage targeting cognitive load theory
notice_stage <- bid_notice(
  problem = "Users overwhelmed by filtering options and dropdown menus",
  theory = "Cognitive Load Theory",
  evidence = "Feedback shows 65% of users abandon the dashboard after first use"
)

notice_stage
notice_stage$theory
notice_stage$suggestions
```
:::
:::

@@

::: {.r-stack}
::: {.fragment .fade-out}
```{r}
#| echo: true

# BID chained stages
project_docs <- notice_stage |>
  bid_interpret(
    central_question = "How can we simplify the filtering experience?",
    # More details are better, but the function will offer suggestions if needed
    data_story = list(
      hook = "Dashboard abandonment hurts decision-making",
      context = "Current UI requires too many decisions before seeing results",
      resolution = "Simplify filter UI with smart defaults",
      audience = "Marketing team",
      metrics = c("Time to insight", "Filter usage")
    )
  ) |>
  bid_structure(
    layout = "dual_process",
    concepts = c("Principle of Proximity", "Default Effect")
  )
```
:::

::: {.fragment}
```{r}
#| echo: true

# BID chained stages output
project_docs
project_docs$suggestions
```
:::
:::

@@

::: {.r-stack}
::: {.fragment .fade-out}
```{r}
#| echo: true

# Get component suggestions based on your BID application
ui_suggestions <- bid_suggest_components(project_docs, package = "bslib")
ui_suggestions
```
:::

::: {.fragment .fade-in}
```{r}
#| echo: true

# Component suggestions output
ui_suggestions$component
ui_suggestions$description
ui_suggestions$code_example
```
:::
:::

@@

- Currently, `{bidux}` has helper functions for all 5 stages of the BID framework [CLICK]
- **Notice tab**
  * `bid_notice` let's you begin by documenting user problems
  * Here we document user problems like "overwhelming dropdown menus" leading to high dashboard abandonment
  * Without specifying a theory, the package suggests one and provides actionable recommendations [SWITCH ON PRESENTATION SCREEN]
- **Concepts tab**
  * If you're unfamiliar with the concept or theory, you can use `bid_concepts` to look them up...[CLICK]...[CLICK]...[CLICK]...or search broadly (like "cog" for cognitive concepts)
  * Maybe you want to update your project to use a difference concept (e.g, cognitive load theory) [CLICK]
  * If so, you can go back and apply them directly in your workflow for new recommendations [SWITCH ON PRESENTATION SCREEN]
- **Chained Stages tab**
  * The real power though is piping through the entire framework
  * We start with our noticed problem, interpret the user's needs, then structure the solution...[CLICK]...[CLICK]...all while getting tailored suggestions at each stage [SWITCH ON PRESENTATION SCREEN]
- **Components tab**
  * And finally, depending on the unique needs of your project, `bid_suggest_components` bridges theory to code with UI component recommendations and implementation examples...[CLICK]...[CLICK]
- This is Phase 1 we're launching at ShinyConf today...[CLICK]...check our GitHub repo for updates
:::

@@

:::: {.columns}
::: {.column width="50%" .fragment}
> "The BID framework transformed how we approach our dashboards. What once took weeks of refinement now has clear direction from day one."
> 
> — Maya Chen, SoundPulse Research Director
:::

@@

### Jeremy Winget, PhD {style="color: white;"}
![](img/headshot-winget.jpg){width="80px" height="80px" style="border-radius: 50%;"}
![](img/qr-linkedin-winget.svg){width="80px" height="80px"}

![](img/lets-connect.png){width="50%"}
:::
::::

>>>>

---
title: "Death By Dropdown?"
subtitle: "A Developer's Guide To Building Dashboards That Won't Fry Your Client's Brain"
author: "Milena Eickhoff & Jeremy Winget, PhD"
format: 
  revealjs:
    title-slide-attributes:
      data-background-image: "img/title.png"
      data-background-size: "100% 100%"
    theme: simple
    slide-number: false
    footer: "ShinyConf 2025 | Death By Dropdown?"
    incremental: true
    transition: slide
    background-transition: fade
    highlight-style: github
    code-block-background: true
execute: 
  echo: true
---

@@

::: {.notes}
- Have you ever felt like this? You were asked to develop a dashboard with endless dropdown menus, overwhelming complexity, that you know would leave users disengaged and searching for answers? 
-	Take a second to respond to our poll in the chat if you have felt like this
  * [Take a look if you want to engage; Jeremy will publish this one as soon as the session starts and message in the chat about it]
  * How many of you have even abandoned a dashboard because it was too complex?
:::

@@

- The Behavior Insight Design framework flows through the 5 stages shown here: Notice → Interpret → Structure → Anticipate → Validate
- [CLICK] **Notice the problem** identifies friction points where users struggle with the interface
- [CLICK] **Interpret the User's Need** uncovers what story users truly need - moving from symptoms to diagnosis
- [CLICK] **Structure the Dashboard** organizes information in ways that align with how our brains process information
- [CLICK] **Anticipate User Behavior** considers cognitive biases affecting data interpretation
- [CLICK] And **Validate & Empower the User** ensures users leave with clear takeaways and a positive impression
- As you can see, the Behavior Insight Design framework is designed to be flexible...you can move between stages as your understanding evolves
:::

@@

::: {.column  .nonincremental .fragment}
- **UX Design**: Established best practices
- **Data Storytelling**: From data dumps to meaningful narratives
- **Psychological Science**: Evidence-based cognitive principles
- **Visual Communication**: Optimized for information perception
:::

@@

- The Behavior Insight Design framework integrates established best practices from multiple disciplines - we're truly standing on the shoulders of giants
  * [CLICK]...UX, data storytelling, psych science, data/visual communication
- What makes BID uniquely powerful is how it systematically connects psychological principles to specific coding decisions at each development stage
- Unlike typical checklists, the Behavior Insight Design framework guides how you think about dashboards from psychological first principles
- The framework bridges theory and practice - showing when and how to apply principles in your workflow
- You don't need a psychology degree to apply these effectively
- Let's see how these stages transformed SoundPulse's dashboard
:::

@@

- **Stage 1 (Notice):** [CLICK] Users were overwhelmed with options
  * [CLICK] This relates to 'cognitive load' - the mental effort needed to process information
  * When the brain gets overtaxed, comprehension drops
  * [CLICK] We also relied on Hick's Law here, which shows decision time increases logarithmically with more choices
- **Stage 2 (Interpret):** [CLICK] We used these concepts in client discussions to help us identify what users needed - increased Processing Fluency [CLICK]
  * When information is easy to process, people find it more credible
  * And we leveraged this to focused on the key story [CLICK] executives needed...not just showing all possible data and letting users make their own choices
- **Stage 3 (Structure):** [CLICK] We applied the 'Principle of Proximity' in the structure stage [CLICK]...which is about grouping related elements together
  * [CLICK] And we made use of Dual-Processing Theory...which distinguishes between fast, intuitive thinking and slower, deliberative thinking...in the design to create logical groupings of metrics for more intuitive workflows
- Remember, the specific concepts can vary...it's the framework's stages that provide consistent guidance
- So...let's take a look at how we used the framework to apply these concepts in our app
:::

@@

```{r}
#| eval: false
#| code-line-numbers: "4-5|6-18|19-23"

@@

```{r}
#| eval: false
#| code-line-numbers: "2-10|12-16|17-27|29-38|39|40-56"

@@

- In the more advanced example we will show you how you can group columns and create a dynamic legend.
  * First, we create a vector that stores the colors we want to use to distinguish the subgroups. 
  * Then we create the reactable. You can use the columnGroups   option to group primary and secondary metrics together (example of the principle of proximity) – it creates a header above the columns. 
  * Below we create a container for the dynamic legend and map over the colors and age groups using a function from the purrr package. The dynamic legend reduces choice complexity for the user.
- "We have another poll in the chat...How many have used columnGroups in reactable before?"
  * [Take a look if you want to engage; Jeremy will publish this one 1-2 slides earlier]
  * If not many…great, we taught something!
  * If a lot…could also mention (quickly) how to extend with legend/div
:::

@@

- For the later Behavior Insight Design stages, we addressed subtler psychological factors affecting interpretation and action
- **Stage 4 (Anticipate):** We expected some folks to make flawed comparisons without proper context
  * This relates to the 'Anchoring Effect'...where people rely too heavily on the first piece of information they see
  * We also saw teams struggling to interpret the data in the same way, which connects to 'Framing & Loss Aversion'...how the presentation of identical data as a "gain" or a "loss" significantly affects decisions
- **Stage 5 (Validate):** Finally, we applied the 'Peak-End Rule'...the idea that people judge experiences by their most intense moment and ending
  * This helped us understand why users needed clear, memorable takeaways to conclude their dashboard experience
- These psychological principles guided our approach to the remaining implementation
:::

@@

```{r}
#| eval: false
#| code-line-numbers: "2-21|23-30"

@@

```{r}
#| eval: false
#| code-line-numbers: "2-21|23-29"

@@

```{r}
#| eval: false
#| code-line-numbers: "6-12|24|25-40|42-57"

@@

  # Generate all chart outputs dynamically with purrr::walk()
  kpi_col_names <- c("popularity", "danceability", "energy")
  walk(kpi_col_names, function(kpi_col) {
    output_id <- paste0("chart_", kpi_col)

@@

- For the framing toggle we use radio buttons that trigger the server-side code [CLICK] to dynamically recalculate the values based on the framing perspective [CLICK]. Different teams at SoundPulse used different frames - marketing used Progress, product used Gap. This is psychological design directly impacting business decisions . 
  * In the last part we loop [CLICK] over the different kpi columns (using a function from the purrr package) and update all charts dynamically with the function from the previous page.
:::

@@

- We're not just sharing theories - we're giving you practical tools for immediate implementation
- The `{bidux}` package directly implements the Behavior Insight Design framework
- Each function maps to a specific stage and guides your thinking
- Phase 1 is available today on GitHub with the conceptual foundation
- Future phases will include LLM integration for suggesting implementations
- We're also developing pre-built UI components embodying these principles
- Think of `{bidux}` as your behavioral science assistant for Shiny development
:::

@@

::: {.fragment}
```{r}
library(bidux)

@@

notice_stage$theory
notice_stage$suggestions
```
:::

@@

::: {.r-stack}
::: {.fragment .fade-out}
```{r}
# Concept dictionary with fuzzy searching and case-insensitive lookup
bid_concept("processing fluency")
```
:::

::: {.fragment .fade-in-then-out}
```{r}
# Concept dictionary with fuzzy searching and case-insensitive lookup
bid_concept("processing fluency")$description
bid_concept("processing fluency")$implementation_tips
```
:::

::: {.fragment .fade-in-then-out}
```{r}
# Concept dictionary with fuzzy searching and case-insensitive lookup
cog_concepts <- bid_concepts("cog")
paste("Found", nrow(cog_concepts), "cog-related concepts")
```
:::

::: {.fragment}
```{r}
# BID Notice stage targeting cognitive load theory
notice_stage <- bid_notice(
  problem = "Users overwhelmed by filtering options and dropdown menus",
  theory = "Cognitive Load Theory",
  evidence = "Feedback shows 65% of users abandon the dashboard after first use"
)

notice_stage$theory
notice_stage$suggestions
```
:::
:::

@@

::: {.r-stack}
::: {.fragment .fade-out}
```{r}
# BID chained stages
project_docs <- notice_stage |>
  bid_interpret(
    central_question = "How can we simplify the filtering experience?",
    # More details are better, but the function will offer suggestions if needed
    data_story = list(
      hook = "Dashboard abandonment hurts decision-making",
      context = "Current UI requires too many decisions before seeing results",
      resolution = "Simplify filter UI with smart defaults",
      audience = "Marketing team",
      metrics = c("Time to insight", "Filter usage")
    )
  ) |>
  bid_structure(
    layout = "dual_process",
    concepts = c("Principle of Proximity", "Default Effect")
  )
```
:::

::: {.fragment}
```{r}
# BID chained stages output
project_docs
project_docs$suggestions
```
:::
:::

@@

::: {.r-stack}
::: {.fragment .fade-out}
```{r}
# Get component suggestions based on your BID application
ui_suggestions <- bid_suggest_components(project_docs, package = "bslib")
ui_suggestions
```
:::

::: {.fragment .fade-in}
```{r}
# Component suggestions output
ui_suggestions$component
ui_suggestions$description
ui_suggestions$code_example
```
:::
:::

@@

- Currently, `{bidux}` has helper functions for all 5 stages of the Behavior Insight Design framework [CLICK]
- **Notice tab**
  * `bid_notice` let's you begin by documenting user problems
  * Here we document user problems like "overwhelming dropdown menus" leading to high dashboard abandonment
  * Without specifying a theory, the package suggests one and provides actionable recommendations [SWITCH ON PRESENTATION SCREEN]
- **Concepts tab**
  * If you're unfamiliar with the concept or theory, you can use `bid_concepts` to look them up...[CLICK]...[CLICK]...[CLICK]...or search broadly (like "cog" for cognitive concepts)
  * Maybe you want to update your project to use a difference concept (e.g, cognitive load theory) [CLICK]
  * If so, you can go back and apply them directly in your workflow for new recommendations [SWITCH ON PRESENTATION SCREEN]
- **Chained Stages tab**
  * The real power though is piping through the entire framework
  * We start with our noticed problem, interpret the user's needs, then structure the solution...[CLICK]...[CLICK]...all while getting tailored suggestions at each stage [SWITCH ON PRESENTATION SCREEN]
- **Components tab**
  * And finally, depending on the unique needs of your project, `bid_suggest_components` bridges theory to code with UI component recommendations and implementation examples...[CLICK]...[CLICK]
- This is Phase 1 we're launching at ShinyConf today...[CLICK]...check our GitHub repo for updates
:::

@@

:::: {.columns}
::: {.column width="50%" .fragment}
<blockquote style="color: #ffca90;">
  <p>"The BID framework transformed how we approach our dashboards. What once took weeks of refinement now has clear direction from day one."</p>
  <p>— Maya Chen, SoundPulse Research Director</p>
</blockquote>
:::

@@

### Jeremy Winget, PhD {style="color: white; margin-top: 10px;"}
![](img/headshot-winget.jpg){width="80px" height="80px" style="border-radius: 50%;"}
![](img/qr-linkedin-winget.svg){width="80px" height="80px"}

![](img/lets-connect.png){width="50%" margin-top="10px"}

<span style="color:white">Slides available at</span> <a href="https://github.com/jrwinget/bid-framework/tree/main/conference-materials/2025_shiny-conf" style="color: #ff9000;">our GitHub repo</a>
:::
::::
